\propertitle

### 2019-06-14 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-13T16:07:01+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The idea behind the approach is easily summarized:
\begin{enumerate*}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full population using the maximum-entropy method;
  \item\label{item:sample_step} the expectation constraints used in the
  method for the full population are, in turn, determined via sampling
  theory from the expectation constraints for the sample.
\end{enumerate*}

Regarding point~\ref{item:maxent_step}, we assume you to be familiar with
the maximum-entropy method. In short, it amounts to a pair of
prescriptions: choose the distribution having maximum Shannon entropy among
those satisfying specific convex constraints, such as expectation values;
and judge these expectations to be equal to measured averages. We add two
remarks that you may not be familiar with. First, the
\enquote{distribution} given by this method is the zeroth-order
approximation of four different distributions:
\begin{itemize}
\item the \emph{probability} distribution for the value of the activity at any
  time bin among those \emph{recorded},
\item the \emph{probability} distribution for the value of the activity at
  a \emph{new} time bin,
\item the most probable \emph{time-frequency} distribution of the activity
  during the \emph{recording},
\item the most probable \emph{time-frequency} distribution of the activity
  in a very long run of \emph{new} time bins.
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
conceptually different inferences. Note that although the four
distributions above numerically coincide in this approximation, they
numerically differ in higher-order approximations. Second, the
maximum-entropy method based on the Shannon entropy implicitly makes
specific assumptions about the probability of the long-run frequency
distributions \citep{jaynes1986d_r1996,portamana2009,portamana2017}. We
discuss these assumptions in \sect***

In our case, to apply the method to the total activity of the full
population we need to fix some of its expectation values, for example its
moments. But we don't have any measured averages for the full population to
equate our expectations to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact relation between the first $m$ moments
for the full population and the first $m$ for the sample; the ones
determine the others and vice versa at every time bin. This is a classical
result of sampling theory \citep[for
example][\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}. Combining this result with the maximum-entropy
prescription \enquote{expectations${}={}$measured averages} for the sample,
we have that the measured averages for the sample determine the moments for
the full population:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured averages} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured averages} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{full-population moments}}_{\textit{sampling theory}}
\end{equation*}

This step is even more straightforward if instead of power moments we use
\emph{normalized factorial moments}. The $m$th normalized factorial moment
is defined as the expectation
\begin{equation}
  \E\condt[\Bigg]{
    \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pf(\ya \| \yH),
\end{equation}
where $\ya\in\set{0, 1/n, 2/n, \dotsc, (n-1)/n, 1}$ is the normalized total
activity of a population of $n$ neurons, and $\pf(\ya \| \yH)$ is the
probability for activity $\ya$ given assumptions $\yH$. This moment can be
interpreted as the expectation of the number of distinct $m$-tuples of
simultaneously active neurons, normalized by the maximum possible number of
distinct $m$-tuples. Note that the first $m$ factorial moments determine
the first $m$ moments and vice versa, so fixing the ones is equivalent to
fixing the others. Now assume that these $n$ neurons are a sample from a
full population of $N$ neurons, with possible normalized activity
$\yA \in\set{0,1/N,\dotsc,1}$. We have this extremely convenient property:
\emph{The first $n-1$ normalized factorial moments for the sample and for
  the full population are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad m < n.
\end{equation}

Hence we can apply the maximum-entropy method to obtain a distribution for
the full population by constraining, for any $m$s we please, its $m$th
factorial moment to be equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$. If $\ya_{t}$ is the normalized
total activity of the sample during the time bin $t$, during a recording
with $T$ time bins, the constraint on the distribution $\pf(\yA \| \yH)$
for the full population is
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA \| \yH)
\end{equation}
for any $m$s we wish.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-14T14:06:05+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The idea behind the approach presented here is easily summarized:
\begin{enumerate}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full population using the maximum-entropy method;
  \item\label{item:sample_step} the expectation constraints used in the
  method for the full population are, in turn, determined via sampling
  theory from the expectation constraints for the sample.
\end{enumerate}
Let's discuss these points in more detail.

Regarding point~\ref{item:maxent_step}, we assume you to be familiar with
the maximum-entropy method. In short, it amounts to a pair of
prescriptions: choose the distribution having maximum Shannon entropy among
those satisfying specific convex constraints, such as expectation values;
and judge these expectation values to be equal to measured averages. We add
two remarks that are seldom made in the literature. First, the
\enquote{distribution} given by this method is the zeroth-order
approximation of four different distributions:
\begin{itemize}
\item the \emph{probability} distribution for the value of the activity at any
  time bin among those \emph{recorded},
\item the \emph{probability} distribution for the value of the activity at
  a \emph{new} time bin,
\item the most probable \emph{time-frequency} distribution of the activity
  during the \emph{recording},
\item the most probable \emph{time-frequency} distribution of the activity
  in a very long run of \emph{new} time bins.
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
conceptually different inferences. Note that although the four
distributions above numerically coincide in this approximation, they
numerically differ in higher-order approximations. Second, the
maximum-entropy method based on the Shannon entropy implicitly makes
specific assumptions about the probability of the long-run frequency
distributions \citep{jaynes1986d_r1996,portamana2009,portamana2017}. We
discuss these assumptions in \sect***

In our case, to apply the method to the total activity of the full
population we need to fix some of its expectation values, for example its
moments. But we don't have any measured averages for the full population to
equate our expectations to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact, linear relation between the first $m$
moments for the full population and the first $m$ for the sample; the ones
determine the others and vice versa at every time bin \citep[for
example][\eqns~(16)]{portamanaetal2015}. This is a classical result of
sampling theory \citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

Combining this result with the maximum-entropy prescription
\enquote{expectations${}={}$measured averages} for the sample, we have that
the measured averages for the sample determine the moments for the full
population:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured averages} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured averages} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{full-population moments}}_{\textit{sampling theory}}
\end{equation*}

These two steps are even more straightforward if instead of power moments
we use \emph{normalized factorial moments}. The $m$th normalized factorial
moment is defined as the expectation
\begin{equation}
  \E\condt[\Bigg]{
    \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pf(\ya \| \yH),
\end{equation}
where $\ya\in\set{0, 1/n, 2/n, \dotsc, (n-1)/n, 1}$ is the normalized total
activity of a population of $n$ neurons, and $\pf(\ya \| \yH)$ is the
probability for activity $\ya$ given assumptions $\yH$. This moment can be
interpreted as the expectation of the number of distinct $m$-tuples of
simultaneously active neurons, normalized by the maximum possible number of
distinct $m$-tuples. For example, with $m=2$, if $na=4$ neurons are active
in a population of $n=5$, we have $\tbinom{4}{2}=6$ distinct pairs of
simultaneously active neurons, and if all five neurons were active we would
have a maximum of $\tbinom{5}{2}=10$ distinct pairs; the normalized number
of pairs is therefore $6/10$. Note that the first $m$ factorial moments
linearly determine the first $m$ moments and vice versa, so fixing the ones is
equivalent to fixing the others. Now assume that these $n$ neurons are a
sample from a full population of $N$ neurons, with possible normalized
activity $\yA \in\set{0,1/N,\dotsc,1}$. We have this extremely convenient
property: \emph{The first $n-1$ normalized factorial moments for the sample
  and for the full population are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad m \le n.
\end{equation}

We can hence apply the maximum-entropy method to obtain a distribution for
the full population by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. If $\ya_{t}$ is the
normalized total activity of the sample during the time bin $t$, during a
recording with $T$ time bins, the constraint on the distribution
$\pf(\yA \| \yH)$ for the full population is
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA \| \yH)
\end{equation}
for any $m$s we wish.
